On the power of conditional independence testing \\ under model-X

### 2022-04-26 arxiv-v4/response_to_reviews.tex

We would like to thank the referee and the associate editor for useful comments and for the opportunity to revise our paper. Based on these suggestions, we made several improvements to the revised manuscript. We summarize these next, and then provide point-by-point responses to the comments.

\section{Overview of Revisions}

To sum up the referee's main concerns, (a) the manuscripts's aim are somewhat disconnected from each other and (b) the robustness and methodological portions of the manuscript are weaker than the other portions. We agree with these assessments, and have made revisions to address both of them. Most importantly, \textbf{we restructured the paper around the main goal of studying fundamental power and optimality properties of model-X methods.} This simultaneously addresses the two main concerns by (a) by making the paper's exposition cleaner and more focused and (b) placing more emphasis on our primary contributions while placing less emphasis on our secondary contributions, the latter including the robustness and methodological aspects. In particular:

\begin{itemize}
	\item We updated Sections 1.2 and 1.3 in the introduction to reflect the focus on power and optimality of MX methods.
	\item We rewrote Section 3 to emphasize that the primary goal of the MX(2) $F$-test is a stepping stone to the power analysis of the CRT, while discussing the methodological and robustness implications as secondary consequences of independent interest.
\end{itemize}
\noindent
\textbf{Additionally, we bolstered the methodological aspect of the paper} as follows:
\begin{itemize}
	\item We clarified the relationship between the MX(2) $F$-test, the dCRT, and the GCM test in Section 3.5, including Tables 1 and 2. We discussed that the MX(2) $F$-test may be preferred to the dCRT due to its speed and that the assumptions necessary for the MX(2) $F$-test and GCM test to control Type-I error do not subsume each other. We argue that all three tests have their relative strengths and weaknesses.
	\item We conjectured that the MX(2) $F$-test continues to control the Type-I error even if the machine learning step is done in sample (Section 3.3), and updated the numerical simulation to provide support for this conjecture (Section 3.4). If true, this conjecture would make the MX(2) $F$-test more directly competitive with the dCRT and GCM tests.
\end{itemize}

\noindent We also changed the semiparametric alternative (18) from the less standard $$\bm Y = (\bm X-\mathbb E[\bm X|\bm Z])^T \beta_n + g_n(\bm Z) + \bm \epsilon \quad \text{(previous version)}$$ to the more standard $$\bm Y = \bm X^T \beta_n + g_n(\bm Z) + \bm \epsilon \quad \text{(this version)}.$$

\noindent 
Finally, we added several smaller revisions (primarily in Section 3) to fully address all the referee's comments. We respond point by point to these comments next.

@@

	\begin{quote}
		``In Section 2, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\bm Y| \bm X, \bm Z}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT against a set of semiparametric alternatives (Section 3.1). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section 3.2), which is of independent interest due to its closed form and weaker assumptions (Section 3.3). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section 3.4) and put this section's results into perspective (Section 3.5) before moving on to stating the desired power results in the next section (Section 4).''
	\end{quote}

	\item \textsl{Regarding the robustness of the CRT test, I feel one could argue that assuming exact knowledge of the first two conditional moments of $X | Z$ is not too much weaker than asking for knowledge of the entire distribution. Are there specific examples where completely correct models for the conditional means and variances are known, but not the conditional distribution? In my opinion, a more relevant analysis would quantify the impact of the error in estimating these conditional moments. Also, it might be interesting to consider a version that does not require estimating the second conditional moment, but potentially places a slightly stronger condition on the error in estimating $\mathbb E[Y | Z]$.}\\

	We agree with the referee that assuming exact knowledge of the first two moments of $X|Z$ is still a strong assumption, which we would not necessarily expect to be satisfied in practice. We nevertheless find it comforting that the MX methodologies considered here are not sensitive to misspecification of the higher moments of the assumed $X|Z$. We agree that a stronger kind of robustness analysis would quantify the impact of the error in estimating these conditional moments and/or shift more of the assumptions onto $\mathbb E[Y|Z]$, but these directions are beyond the scope of the current work. We convey these points in Section 3.5 of the revised manuscript:
	\begin{quote}
	``...Theorem 3 shows that asymptotic Type-I error control of MX-style methodologies can be achieved under the weaker MX(2) assumption, requiring only two moments of the conditional distribution $\mathcal L_n(\bm X | \bm Z)$ rather than the entire conditional distribution...note that the MX(2) assumption still requires \textit{exact} knowledge of the first and second conditional moments; we leave as an important future direction to examine the robustness of these tests to errors in these quantities. First steps in this direction have been taken recently (Berrett et al 2020, Li and Liu 2022).''
	\end{quote}

@@

	\item \textsl{What is the significance of the asymptotic equivalence of the CRT test constructed using the test statistic on which the MX(2) $F$-test is based, and the MX(2) $F$-test itself? I cannot see why the CRT test would be preferred here? Overall, the connection between CRT tests and the MX(2) $F$-test seems quite weak to me.} \\

	The significance is that we can study the power of the CRT by instead studying the power of the easier-to-analyze MX(2) $F$-test. We have clarified this motivation to construct the MX(2) $F$-test in the first paragraph of Section 3 (quoted above). Furthermore, we expand on the methodological advantages and disadvantages of the MX(2) $F$-test compared to the CRT in Section 3.5 of the revised manuscript:
	\begin{quote}
		``The preceding results suggest that the MX(2) $F$-test is a useful alternative to the dCRT: the power of these methods is asymptotically the same (Theorem 2), while the MX(2) $F$-test is computationally faster because it does not require resampling (Table~\ref{tab:summary-1}). On the other hand, note that we have proven Type-I error control for the MX(2) $F$-test only when $\widehat g_n$ is fit out of sample and only asymptotically, while the dCRT gives finite-sample Type-I error with in-sample fit $\widehat g_n$ (albeit under the stronger model-X assumption). However, numerical simulations suggest good finite-sample Type-I error control for the MX(2) $F$-test even when $\widehat g_n$ is fit in sample.''
	\end{quote}

### 2022-04-26 arxiv-v4/crtpower.tex
Incorporated Aaditya's suggestions

\usepackage[linesnumbered,ruled]{algorithm2e}

@@

Now, consider a set of semiparametric alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = (\prx-\mu_n(\prz))^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. By Theorem~\ref{prop:crt-optimality}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
T_n^{\text{opt}}(\srx, \sry, \srz) = \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n(\srz_i)\right)^2\right).
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T_n^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - g_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - g_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

Of course, inference based on $T_n^{\text{opt}}$ is infeasible because the function $g_n$ is unknown in practice. Suppose we have an estimate $\widehat g_n$ of this function. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
Consider first the case $d = 1$. In this case, $\widehat \rho_n$ coincides with the recently introduced \textit{generalized covariance measure} (GCM) \cite{Shah2018}; see Section~\ref{sec:comparison-to-existing-results-3} for more a comparison to the latter work. The CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the dCRT \cite{Liu2020}, proposed for its computational speed: If $\widehat g_n$ is fit only on $(\sry, \srz)$, then it need not be refit upon resampling $\widetilde X$. The above reasoning suggests that the dCRT is also a natural test to study in the context of power against semiparametric alternatives~\eqref{eq:linearity}. In preparation to study this test's power, we extend it to $d > 1$ and propose an asymptotically equivalent test that is easier to understand.

@@

Let us consider first the special case
\begin{equation}
\mathcal L_n(\prx | \prz) = N(\mu_n(\prz), \Sigma_n(\prz)).
\label{eq:normal-conditional}
\end{equation}
In this case, the resampling distribution of $\widehat \rho_n$ can be computed in closed form \cite{Liu2020}:
\begin{equation}
\mathcal L_n(\sqrt n \cdot \widehat \rho_n(\srxk, \sry, \srz) \mid \srx, \sry, \srz) = N(0, \widehat S_n^2),
\label{eq:normal-resampling-distribution}
\end{equation}
where
\begin{equation}
	\widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i).
	\label{eq:s-n-hat}
\end{equation}
When $d = 1$, the dCRT based on the statistic $T_n(\srx, \sry, \srz) = |\sqrt n \cdot \widehat \rho_n(\srx, \sry, \srz)|$ therefore rejects when $T_n(\srx, \sry, \srz) > \widehat S_n \cdot z_{1-\alpha/2}$, requiring no resampling. To extend this to $d > 1$, consider the standardized quantity
\begin{equation}
U_n(\srx, \sry, \srz) \equiv \widehat S_n^{-1} \sqrt{n} \widehat \rho_n = \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)) \in \mathbb R^d.
\label{u-hat}
\end{equation}
It is natural to use as a test statistic the squared norm of $U_n$:
\begin{equation}
	T_n(\srx, \sry, \srz) \equiv \|U_n(\srx, \sry, \srz)\|^2.
	\label{t-n}
\end{equation}
Then, the normal resampling distribution~\eqref{eq:normal-resampling-distribution} implies that
\begin{equation}
\mathcal L_n(T_n(\srxk, \sry, \srz)|\srx, \sry, \srz) = \chi^2_d.
\end{equation}
It follows that the dCRT based on test statistic $T_n(\srx, \sry, \srz)$ yields the test
\begin{equation}
\phi^{N(\mu_n, \Sigma_n)}_n(\srx, \sry, \srz) \equiv \mathbbm 1(T_n(\srx, \sry, \srz) > c_{d,1-\alpha}),
\label{eq:crt-closed-form}
\end{equation}
where we recall that $c_{d,1-\alpha}$ is defined as the $1-\alpha$ quantile of $\chi^2_d$. Note that all tests $\phi$ in Sections~\ref{sec:weakening} and~\ref{sec:asymptotic-power} will be (d)CRTs based on the test statistic $T_n$~\eqref{t-n}. To ease notation, we therefore omit the subscript $T_n$ and the superscript ``CRT'' from the notation introduced in equation~\eqref{eq:randomized-CRT}, replacing these with $n$ and the distribution of $\prx|\prz$ with respect to which resampling is done, respectively. For example, the superscript in the test defined in equation~\eqref{eq:crt-closed-form} is based on the resampling distribution $\prx|\prz \sim N(\mu_n(\prz), \Sigma_n(\prz))$.

@@

\begin{theorem} \label{thm:equivalence}
Suppose that for each $n$, $\mathcal L_n$ is a law whose first and second conditional moments are given by $\mu_n(\prz)$ and $\Sigma_n(\prz)$~\eqref{eq:conditional-mean-variance}, which satisfies the moment conditions~\eqref{eq:moment-conditions} for fixed for some $c_1, c_2 > 0$. Let $\phi_n^{\mathcal L_n}$ be the dCRT based on the test statistic $T_n(\srx, \sry, \srz)$~\eqref{eq:s-n-hat}, \eqref{u-hat}, \eqref{t-n}, with $\widehat g_n$ trained out of sample. The threshold $C_n(\sry, \srz)$ of this test~\eqref{upper-quantile} converges in probability to the $\chi^2_d$ quantile:
	\begin{equation}
		C_n(Y,Z) \overset{\mathcal L_n}\rightarrow_p c_{d,1-\alpha}.
		\label{eq:threshold-convergence}
	\end{equation}
	Furthermore, if $T_n(\srx, \sry, \srz)$ does not accumulate near $c_{d,1-\alpha}$, i.e.
	\begin{equation}
		\lim_{\delta \rightarrow 0}\limsup_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[|T_n(\srx, \sry, \srz)-c_{d,1-\alpha}| \leq \delta] = 0,
		\label{eq:non-accumulation}
	\end{equation}
	then the dCRT $\phi_n^{\mathcal L_n}$ is asymptotically equivalent to its second order approximation $\phi^{N(\mu_n, \Sigma_n)}_n$:
	\begin{equation}
		\lim_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[\phi^{\mathcal L_n}_n(\srx, \sry, \srz) \neq \phi^{N(\mu_n, \Sigma_n)}_n(\srx, \sry, \srz)] = 0.
		\label{eq:asymptotic-equivalence}
	\end{equation}	
\end{theorem}

@@

Under the MX(2) assumption, the CRT is undefined because there is no conditional distribution $\mathcal L_n(\prx|\prz)$ to resample from. Nevertheless, we may define the \textit{MX(2) $F$-test} by running the resampling-free dCRT as though $\mathcal L_n(\prx|\prz)$ were normal, with the given first and second moments (Algorithm~\ref{alg:MX(2)-F-test}). We denote this test $\phi_n^{N(\mu_n, \Sigma_n)}$, as before.
\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\KwData{
				$\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, $\mu_n(\cdot)$ and $\Sigma_n(\cdot)$ in \eqref{eq:conditional-mean-variance}, learning method $g$
			}
			Obtain $\widehat g_n$ by fitting $g$ (either in-sample, out-of-sample, or a mix of both)\;
			Recall $\mu_n(\srz_i) \equiv \mathbb E_{\mathcal L_n}[\srx_i|\srz_i], \text{ set } \widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i)$\;
			Set $\smash{U_n \equiv \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i))}$ and $T_n = \|U_n\|^2$\;
			% Compute $T(\srx,\sry,\srz) \equiv \sum_{i = 1}^n \ell(\sry_i, (\srx_i-\mu_n(\srz_i)) \widehat \beta + \widehat g_n(\srz_i))$\;
			% \For{$b \in \{1,\dots,B\}$}{
			% 	Sample $\srxk_i^b$ from $\mathcal L(\srx_i|\srz_i)$ for all $i$\;
			% 	Compute $T(\srxk^b,\sry,\srz)$ using steps 3, 4 with $\srxk_i^b$ replacing $\srx_i$\;
			% }
			% Compute $\widehat p_j$ based on formula~\eqref{CRT}\;
			\KwResult{MX(2) $F$-test asymptotic $p$-value $\widehat p \equiv \mathbb P[\chi^2_d > T_n]$.}
			\textbf{Cost:} One $p$-dimensional model fit.
			\caption{\bf The MX(2) $\bm F$-test}
			\label{alg:MX(2)-F-test}
		\end{algorithm}
	\end{minipage}
\end{center}
Note that a one-sided version of this test (the \textit{MX(2) $t$-test}) can be defined for $d = 1$ by rejecting for large values of $U_n(\srx, \sry, \srz)$.

@@

The preceding results suggest that the MX(2) $F$-test is a useful alternative to the dCRT: the power of these methods is asymptotically the same (Theorem~\ref{thm:equivalence}), while the MX(2) $F$-test is computationally faster because it does not require resampling (Table~\ref{tab:summary-1}). On the other hand, note that we have proven Type-I error control for the MX(2) $F$-test only when $\widehat g_n$ is fit out of sample and only asymptotically, while the dCRT gives finite-sample Type-I error control with in-sample fit $\widehat g_n$. However, numerical simulations suggest good finite-sample Type-I error control for the MX(2) $F$-test even when $\widehat g_n$ is fit in sample. Furthermore, Theorem~\ref{thm:asymptotic-alpha-level} shows that asymptotic Type-I error control of MX-style methodologies can be achieved under the weaker MX(2) assumption~\eqref{eq:mx-2-assumption}, requiring only two moments of the conditional distribution $\mathcal L_n(\prx|\prz)$ rather than the entire conditional distribution (Table~\ref{tab:summary-2}). If strict Type-I error control in finite samples is desired, however, then we must continue to rely on the full MX assumption. %Nevertheless, our simulation suggests the MX(2) $F$-test has fairly good Type-I error control in finite samples as well. 
Finally, note that the MX(2) assumption still requires \textit{exact} knowledge of the first and second conditional moments; we leave as an important future direction to examine the robustness of these tests to errors in these quantities. Note that bounds on the worst-case Type-I error of the CRT have been obtained in the context of a misspecified $\mathcal L_n(\prx|\prz)$ \cite{Berrett2019}.

@@

		&=\mathbb E_{\mathcal L_n}[( \peps + g_n(\prz)-\widehat g_n(\prz))^2 \Sigma_n(\prz)] \\

>>>>

\usepackage[linesnumbered,ruled]{algorithm2e}
\let\oldnl\nl% Store \nl in \oldnl
\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line
\makeatother

@@

Now, consider a set of semiparametric alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = \prx^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. By Theorem~\ref{prop:crt-optimality}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
\begin{split}
T_n^{\text{opt}}(\srx, \sry, \srz) &= \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - \srx_i^T \beta_n - g_n(\srz_i)\right)^2\right) \\
&=\prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n'(\srz_i)\right)^2\right),
\label{eq:likelihood-ratio}
\end{split}
\end{equation}
where
\begin{equation}
g'_n(\prz) \equiv \mathbb E[\pry|\prz] =  \mu_n(\prz)^T\beta_n  + g_n(\prz).
\label{eq:g-n-prime-def}
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T_n^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - g'_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - g'_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

Of course, inference based on $T_n^{\text{opt}}$ is infeasible because the function $g'_n$ is unknown in practice. Suppose we have learned an estimate $\widehat g_n$ of this function, possibly in-sample. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
Consider first the case $d = 1$. The CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the dCRT \cite{Liu2020}. The idea of the dCRT (Algorithm~\ref{alg:dCRT}) is to \textit{distill}---usually via machine learning---the information from the high-dimensional $\srz \in \mathbb R^{p \times n}$ about $\srx$ and $\sry$ into a low-dimensional summary $D \in \mathbb R^{q \times n}$,  where $q \ll p$. This is accomplished using a \textit{distillation function} $d: (\sry, \srz) \mapsto D$. Then, the CRT is applied using a test statistic of the form $T_n(\srx, \sry, \srz) \equiv T_n^d(\srx, \sry, D) = T_n^d(\srx, \sry, d(\sry, \srz))$. For example, the CRT based on the statistic $\widehat \rho_n$~\eqref{rho-hat} can be expressed as the dCRT with distillation function $d_i(\sry, \srz) = (\widehat g_n(\srz_i), \mu_n(Z_i))$, where $\widehat g_n$ is learned in-sample on $(\sry, \srz)$.

\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\nonl  \textbf{Input:} $\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, distribution  $f^*_{\prx|\prz}$, distillation function $d$, test statistic $T_n^d$, number of resamples $B$\\
			Distill information in $\srz$ about $\srx$ and $\sry$ into $D \equiv d(\sry, \srz)$\;
			\For{$b = 1, 2, \dots, B$}{
				Resample $\srxk_i^{(b)} \overset{\text{ind}}\sim f^*_{\prx|\prz = \srz_i},\ i = 1, \dots, n$\;
			}
			Compute $\widehat p \equiv \frac{1}{B+1}\sum_{b = 1}^B \mathbbm 1(T_n^d(\srxk^{(b)}, \sry, D) \geq T_n^d(\srx, \sry, D)$.\\
			\nonl \textbf{Output:} dCRT $p$-value $\widehat p$.\\
			\nonl \textbf{Computational cost:} One $p$-dimensional model fit, and drawing $B$ resamples.
			\caption{\bf The distilled conditional randomization test}
			\label{alg:dCRT}
		\end{algorithm}
	\end{minipage}
\end{center}

The dCRT was proposed for its computational speed: The computationally expensive distillation step is a function only of $(\sry, \srz)$, so it need not be refit upon resampling $\widetilde X$. By contrast, the originally proposed instance of the CRT \cite{CetL16} involved learning $\widehat f_{\pry|\prx, \prz}$ on the entire sample $(\srx, \sry, \srz)$, and therefore the learning procedure needed to be re-applied to each resampled dataset $(\srxk^{(b)}, \sry, \srz)$. The derivations~\eqref{eq:likelihood-ratio} and~\eqref{eq:optimal-semiparametric} suggest that the dCRT is not only computationally fast, but also a natural test to consider for power against semiparametric alternatives~\eqref{eq:linearity}. We therefore focus on this class of tests.

In preparation to study the power of the dCRT, we extend it to $d > 1$ and propose an asymptotically equivalent test that is easier to analyze.

% In this case, $\widehat \rho_n$ coincides with the recently introduced \textit{generalized covariance measure} (GCM) \cite{Shah2018}; see Section~\ref{sec:comparison-to-existing-results-3} for more a comparison to the latter work.

@@

Let us consider first the special case
\begin{equation}
\mathcal L_n(\prx | \prz) = N(\mu_n(\prz), \Sigma_n(\prz)).
\label{eq:normal-conditional}
\end{equation}
In this case, the resampling distribution of $\widehat \rho_n$ can be computed in closed form \cite{Liu2020}:
\begin{equation}
\mathcal L_n(\sqrt n \cdot \widehat \rho_n(\srxk, \sry, \srz) \mid \srx, \sry, \srz) = N(0, \widehat S_n^2),
\label{eq:normal-resampling-distribution}
\end{equation}
where
\begin{equation}
	\widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i).
	\label{eq:s-n-hat}
\end{equation}
When $d = 1$, the dCRT based on the statistic $T_n(\srx, \sry, \srz) = |\sqrt n \cdot \widehat \rho_n(\srx, \sry, \srz)|$ (and infinitely many resamples $B$) therefore rejects when $T_n(\srx, \sry, \srz) > \widehat S_n \cdot z_{1-\alpha/2}$, requiring no resampling. To extend this to $d > 1$, consider the standardized quantity
\begin{equation}
U_n(\srx, \sry, \srz) \equiv \widehat S_n^{-1} \sqrt{n} \widehat \rho_n = \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)) \in \mathbb R^d.
\label{u-hat}
\end{equation}
It is natural to use as a test statistic the squared norm of $U_n$:
\begin{equation}
	T_n(\srx, \sry, \srz) \equiv \|U_n(\srx, \sry, \srz)\|^2.
	\label{t-n}
\end{equation}
Then, the normal resampling distribution~\eqref{eq:normal-resampling-distribution} implies that
\begin{equation}
\mathcal L_n(T_n(\srxk, \sry, \srz)|\srx, \sry, \srz) = \chi^2_d.
\end{equation}
It follows that the dCRT based on test statistic $T_n(\srx, \sry, \srz)$ yields the test
\begin{equation}
\phi^{N(\mu_n, \Sigma_n)}_n(\srx, \sry, \srz) \equiv \mathbbm 1(T_n(\srx, \sry, \srz) > c_{d,1-\alpha}),
\label{eq:crt-closed-form}
\end{equation}
where we recall that $c_{d,1-\alpha}$ is defined as the $1-\alpha$ quantile of $\chi^2_d$. Note that all tests $\phi$ in Sections~\ref{sec:weakening} and~\ref{sec:asymptotic-power} will be (d)CRTs based on the test statistic $T_n$~\eqref{t-n}. To ease notation, we therefore omit the subscript $T_n$ and the superscript ``CRT'' from the notation introduced in equation~\eqref{eq:randomized-CRT}, replacing these with $n$ and the distribution of $\prx|\prz$ with respect to which resampling is done, respectively. For example, the superscript in the test defined in equation~\eqref{eq:crt-closed-form} is based on the resampling distribution $\prx|\prz \sim N(\mu_n(\prz), \Sigma_n(\prz))$.

@@

\begin{theorem} \label{thm:equivalence}
Suppose that for each $n$, $\mathcal L_n$ is a law whose first and second conditional moments are given by $\mu_n(\prz)$ and $\Sigma_n(\prz)$~\eqref{eq:conditional-mean-variance}, which satisfies the moment conditions~\eqref{eq:moment-conditions} for fixed for some $c_1, c_2 > 0$. Let $\phi_n^{\mathcal L_n}$ be the dCRT based on the test statistic $T_n(\srx, \sry, \srz)$~\eqref{eq:s-n-hat}, \eqref{u-hat}, \eqref{t-n}, with $\widehat g_n$ trained out of sample. The threshold $C_n(\sry, \srz)$ of this test~\eqref{upper-quantile} converges in probability to the $\chi^2_d$ quantile:
	\begin{equation}
		C_n(Y,Z) \overset{\mathcal L_n}\rightarrow_p c_{d,1-\alpha}.
		\label{eq:threshold-convergence}
	\end{equation}
	Furthermore, if $T_n(\srx, \sry, \srz)$ does not accumulate near $c_{d,1-\alpha}$, i.e.
	\begin{equation}
		\lim_{\delta \rightarrow 0}\limsup_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[|T_n(\srx, \sry, \srz)-c_{d,1-\alpha}| \leq \delta] = 0,
		\label{eq:non-accumulation}
	\end{equation}
	then the dCRT $\phi_n^{\mathcal L_n}$ is asymptotically equivalent to its second order approximation $\phi^{N(\mu_n, \Sigma_n)}_n$~\eqref{eq:crt-closed-form}:
	\begin{equation}
		\lim_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[\phi^{\mathcal L_n}_n(\srx, \sry, \srz) \neq \phi^{N(\mu_n, \Sigma_n)}_n(\srx, \sry, \srz)] = 0.
		\label{eq:asymptotic-equivalence}
	\end{equation}	
\end{theorem}

@@

Under the MX(2) assumption, the CRT is undefined because there is no conditional distribution $\mathcal L_n(\prx|\prz)$ to resample from. Nevertheless, we may define the \textit{MX(2) $F$-test} by running the resampling-free dCRT as though $\mathcal L_n(\prx|\prz)$ were normal, with the given first and second moments (Algorithm~\ref{alg:MX(2)-F-test}). We denote this test $\phi_n^{N(\mu_n, \Sigma_n)}$, as before.
\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\KwData{
				$\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, $\mu_n(\cdot)$ and $\Sigma_n(\cdot)$ in \eqref{eq:conditional-mean-variance}, learning method $g$
			}
			Obtain $\widehat g_n$ by fitting $g$ (either in-sample, out-of-sample, or a mix of both)\;
			Recall $\mu_n(\srz_i) \equiv \mathbb E_{\mathcal L_n}[\srx_i|\srz_i], \text{ set } \widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i)$\;
			Set $\smash{U_n \equiv \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i))}$ and $T_n = \|U_n\|^2$\;
			% Compute $T(\srx,\sry,\srz) \equiv \sum_{i = 1}^n \ell(\sry_i, (\srx_i-\mu_n(\srz_i)) \widehat \beta + \widehat g_n(\srz_i))$\;
			% \For{$b \in \{1,\dots,B\}$}{
			% 	Sample $\srxk_i^b$ from $\mathcal L(\srx_i|\srz_i)$ for all $i$\;
			% 	Compute $T(\srxk^b,\sry,\srz)$ using steps 3, 4 with $\srxk_i^b$ replacing $\srx_i$\;
			% }
			% Compute $\widehat p_j$ based on formula~\eqref{CRT}\;
			Compute $\widehat p^{\text{MX(2)}} \equiv \mathbb P[\chi^2_d > T_n]$.\\
			\nonl \textbf{Output:} MX(2) $F$-test asymptotic $p$-value $\widehat p^{\text{MX(2)}}$.\\
			\nonl \textbf{Computational cost:} One $p$-dimensional model fit.
			\caption{\bf The MX(2) $\bm F$-test}
			\label{alg:MX(2)-F-test}
		\end{algorithm}
	\end{minipage}
\end{center}
Note that a one-sided version of this test (the \textit{MX(2) $t$-test}) can be defined for $d = 1$ by rejecting for large values of $U_n(\srx, \sry, \srz)$.

@@

The preceding results suggest that the MX(2) $F$-test is a useful alternative to the dCRT: the power of these methods is asymptotically the same (Theorem~\ref{thm:equivalence}), while the MX(2) $F$-test is computationally faster because it does not require resampling (Table~\ref{tab:summary-1}). On the other hand, note that we have proven Type-I error control for the MX(2) $F$-test only when $\widehat g_n$ is fit out of sample and only asymptotically, while the dCRT gives finite-sample Type-I error control with in-sample fit $\widehat g_n$ (albeit under the stronger model-X assumption). However, numerical simulations suggest good finite-sample Type-I error control for the MX(2) $F$-test even when $\widehat g_n$ is fit in sample. Furthermore, Theorem~\ref{thm:asymptotic-alpha-level} shows that asymptotic Type-I error control of MX-style methodologies can be achieved under the weaker MX(2) assumption~\eqref{eq:mx-2-assumption}, requiring only two moments of the conditional distribution $\mathcal L_n(\prx|\prz)$ rather than the entire conditional distribution (Table~\ref{tab:summary-2}). If strict Type-I error control in finite samples is desired, however, then we must continue to rely on the full MX assumption. %Nevertheless, our simulation suggests the MX(2) $F$-test has fairly good Type-I error control in finite samples as well. 
Finally, note that the MX(2) assumption still requires \textit{exact} knowledge of the first and second conditional moments; we leave as an important future direction to examine the robustness of these tests to errors in these quantities. First steps in this direction have been taken recently \cite{Berrett2019, Li2022a}.

@@

		&=\mathbb E_{\mathcal L_n}[( \peps + g'_n(\prz)-\widehat g_n(\prz))^2 \Sigma_n(\prz)] \\
